Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-11-05 paper/paper.tex
Tighten intro paragraphs in Section 2; add an absurd eliminator (still need to add to the proofs!)

\let\Bbbk\relax % amsmath tries to redefine this
\usepackage{amsthm,amssymb}
\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}
\usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
\usepackage{xspace,stmaryrd,bbding}
\usepackage[para]{footmisc}
\usepackage{ottalt}
\inputott{rules}

@@

\begin{figure}[H]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
\end{array}
\]
\[
\begin{array}{lll}
\mathit{Substitutions}      & \qquad\qquad & \mbox{\emph{Typing Contexts}}  \\
[[rho]] \in [[SVar -> STm]] & & [[G]] ::= [[empty]]\ |\ [[G ++ x : A]] 
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

In this section, we present the dependent type theory, \lang{}, whose logical
consistency will be proven in Section~\ref{sec:logrelproof}.

\scw{Need to acknowledge the lack of type annotation on abstractions.}

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent a universe type, annotated by its
universe level, a natural number $i$.  Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
  bound variables are distinct.  In some places, we are informal about the
  treatment of variables and substitution; our mechanized proofs make these
  notions precise by using de Bruijn indices~\cite{debruijn1994automath}. }
We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
on the input variable. For simplicity, we omit the type annotations in the abstraction
forms. We discuss how the inclusion of type annotations can affect
our development in Section~\ref{sec:betaeta}, where we extend our
consistency result to the existence of $\beta\eta$-normal forms.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.  Finally, \lang{}
includes booleans, with standard syntax.  \scw{Example of what you can use
  identity types for? Or an example of a program that uses J?}

Our reduction and typing relations are defined in terms of \emph{simultaneous
  substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
use $[[idtm]]$ as the identity substitution.  The extension operation,
$[[(rho .: x -> a)]]$, updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.

The substitution operator, which takes the form $[[a {rho}]]$, traverses the
syntax of $[[a]]$ and replaces each variable $[[x]]$ with the term
$[[rho x]]$. When traversing under binders (e.g. in the
$[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
bound variable to itself and that the bound variable does not appear freely in
the application of the substitution to any other variable.
%
\scw{Maybe we should handwave more here, and note that we aren't covering all
  of the details of the treatment of variable binding in the text. If readers
  want to understand that part, they should look at the Coq code, which uses
  de Bruijn indices anyways. Our goal is to convey the understanding of the
  Coq proof. }
\yl{I agree. Should mention upfront that the presentation is not
  watertight and rigorous, though we have one in our mechanization}

The substitution operator is referred to as simultaneous substitution as it
substitutes for all variables at once. It is possible to recover single
substitution by composing the extension operator and the identity
substitution: $[[a { b / x }]] := [[a { idtm .: x -> b }]]$.

When reasoning about logical relations, we find it more convenient to
formulate simultaneous substitution directly rather than recovering it from
single substitution.  In particular, this shows up in the definition of
semantic typing in Section~\ref{sec:logrelproof}, which relies on simultaneous
substitution.

@@

\begin{figure}[H]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
% \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
% \drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only) }
\label{fig:par}
\end{figure}

>>>>

\let\Bbbk\relax % amsmath tries to redefine this
\usepackage{amsthm,amssymb}
\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}
\usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
\usepackage{xspace,stmaryrd,mathtools,bbding}
\usepackage[para]{footmisc}
\usepackage{ottalt}
\inputott{rules}

@@

\begin{figure}[h]
\begin{align*}
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & \Coloneqq && \textit{Terms} \\
            & \mid [[Set i]] \mid [[x]] \mid [[Void]] \mid [[absurd b]]
              && \textrm{universes, variables, empty type, explosion} \\
            & \mid [[Pi x : A . B]] \mid [[\ x . a]] \mid [[a b]]
              && \textrm{function types, abstractions, applications} \\
            & \mid [[a ~ b : A ]] \mid [[refl]] \mid [[J c p]]
              && \textrm{equality types, reflexivity proof, J eliminator} \\
            & \mid [[Bool]] \mid [[true]] \mid [[false]] \mid [[if a b0 b1]]
              && \textrm{boolean type, true, false, conditionals} \\
            % & \mid [[if a b0 b1]]
            %   && \textrm{conditional expression} \\
[[G]] & \Coloneqq [[empty]] \mid [[G ++ x : A]] && \textit{Typing contexts} \\
[[rho]] & \in [[SVar -> STm]] && \textit{Substitutions}
\end{align*}
\caption{Syntax of \lang}
\label{fig:syntax}
\end{figure}

In this section, we present the dependent type theory \lang{}, whose logical
consistency will be proven in Section~\ref{sec:logrelproof}.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. We use a Curry-style (extrinsic) syntax, where terms are not
annotated by types, and only terms which participate in reduction are part of
introduction and elimination forms.

The type $[[Set i]]$ represents a universe at level $i$,
a natural number. Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.%
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
  bound variables are distinct.  In some places, we are informal about the
  treatment of variables and substitution; our mechanized proofs make these
  notions precise by using de Bruijn indices~\citep{debruijn1994automath}. }
We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
on the input variable. Type annotations in abstraction forms are omitted, and
we discuss how the inclusion of type annotations can affect
our development in Section~\ref{sec:betaeta}, where we extend our
consistency result to the existence of $\beta\eta$-normal forms.
Intensional identity types $[[a ~ b : A]]$, or equality types, represent an
equality between terms $[[a]]$ and $[[b]]$ of the same type $[[A]]$, whose
canonical inhabitant is the reflexive proof $[[refl]]$. Equality proofs $[[p]]$
can be eliminated by the J eliminator $[[J c p]]$, which casts the term $[[c]]$
from one end of the equality to the other. Finally, we have booleans along with
the standard boolean values and conditional expressions.

Our reduction and typing relations are defined in terms of \emph{simultaneous
substitutions} $[[rho]]$, which are mappings from variables to terms. We
use $[[idtm]]$ as the identity substitution, and the extension operator
$[[rho .: x -> a]]$ updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.
% When traversing under binders (e.g. in the
% $[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
% bound variable to itself and that the bound variable does not appear freely in
% the application of the substitution to any other variable.
%
\scw{Maybe we should handwave more here, and note that we aren't covering all
  of the details of the treatment of variable binding in the text. If readers
  want to understand that part, they should look at the Coq code, which uses
  de Bruijn indices anyways. Our goal is to convey the understanding of the
  Coq proof. }
\yl{I agree. Should mention upfront that the presentation is not
  watertight and rigorous, though we have one in our mechanization}
The substitution operator $[[a {rho}]]$ replaces every free variable $[[x]]$ of
$[[a]]$ by $[[rho x]]$ simultaneously. A single substitution corresponds to
composing extension and identity:
$[[a { b / x }]] \coloneqq [[a { idtm .: x -> b }]]$.

When reasoning about logical relations, it is more convenient to formulate
definitions in terms of simultaneous substitutions than single substitutions,
such as semantic typing in Section~\ref{sec:logrelproof}.
\jc{I've rewritten this paragraph to be more succinct,
but I'm still not certain it's necessary at all.
We're not focussing on the merits of simultaneous substitution over
other representations of substitution and binding, after all.}

@@

\begin{figure}[H]
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\caption{Parallel reduction ($\beta$ rules only) }
\label{fig:par}
\end{figure}
